---
title: "Getting started"
teaching: 30
exercises: 10
questions:
- "FIXME"
objectives: 
- "FIXME"
keypoints:
- "FIXME"
source: Rmd
---

```{r, include=FALSE}
source("../bin/chunk-options.R")
knitr_fig_path("01-")
source("../bin/download_data.R")
library(tidyverse)
library(readxl)
```


## Read in data in R

First we load `tidyverse` which make datamanipulation easier. We also load a 
package to help us read Excel files:
```{r, purl=FALSE}
library(tidyverse)
library(readxl)
```

In order to do useful and interesting things, we need to assign _values_ to
_objects_. To create an object, we need to give it a name followed by the
assignment operator `<-`, and the value we want to give it.

The value in this case, is the result of reading an Excel file:

```{r echo = F}
flightdata <- read_excel("../data/flightdata.xlsx")
```


```{r, purl=FALSE, eval = F}
flightdata <- read_excel("data/flightdata.xlsx")
```


# Taking a look at the data

Always begin by taking a look at your data!

```{r, purl=FALSE, echo = F}
flightdata %>% 
  sample_frac(.005) %>% 
  view()
```

In this case it is the function sample_frac() that takes the input, flightdata, and returns a random fraction of the data. 0.005 in this case. 
The dataset is pretty large. 

The summary function returns summary statistics on our data:
```{r}
summary(flightdata)
```

```{r}
flightdata %>% 
  filter(dep_delay == 0) %>% 
  nrow()
```


What is actually contained in this dataset, what are the meaning of 
the column names?

* year, month, day Date of departure.
* dep_time, arr_time Actual departure and arrival times (format HHMM or HMM), local tz.
* sched_dep_time, sched_arr_time Scheduled departure and arrival times (format HHMM or HMM), local tz.
* dep_delay, arr_delay Departure and arrival delays, in minutes. Negative times represent early departures/arrivals.
* carrier Two letter carrier abbreviation. See airlines to get name.
* flight Flight number.
* tailnum Plane tail number. See planes for additional metadata.
* origin, dest Origin and destination. See airports for additional metadata.
* air_time Amount of time spent in the air, in minutes.
* distance Distance between airports, in miles.
* hour, minute Time of scheduled departure broken into hour and minutes.
* time_hour Scheduled date and hour of the flight as a POSIXct date. Along with origin, can be used to join flights data to weather data.

Always remember to save information about what is actually in your 
data. This is called metadata, data that describes data. 


## Let us make a plot

```{r}
flightdata %>% 
  sample_frac(.005) %>% 
  ggplot(mapping = aes( x = dep_delay, y = arr_delay)) +
  geom_point()
```

Note that the pipe is not a pipe in plots. Its a +.

Mapping denotes which values from the data, should be mapped to 
something in the plot. The "somethings" depends on the type of plot. We are making a scatter plot, using geom_point, because we are plotting points. They have a minimum requirement of x and y.





{% include links.md %}
