---
title: "Starting with Data"
teaching: 30
exercises: 10
questions:
- "What is a data.frame?"
- "How can I read a complete csv file into R?"
- "How can I get basic summary information about my dataset?"
objectives:
- "Describe what a data frame is."
- "Load external data from a .csv file into a data frame."
- "Summarize the contents of a data frame."
- "Subset and extract values from data frames."
keypoints:
- "Use read_csv to read tabular data in R."
source: Rmd
---


```{r, include=FALSE}
source("../bin/chunk-options.R")
knitr_fig_path("02-")
library(tidyverse)
source("../bin/download_data.R")
```


Endnu tættere blik på data - boxplots.

histogrammer

facetter

density curves - er der forskelle på de tre lufthavne?

Joins

Er der forskel på forsinkelser som funktion af flyenes alder?

lineær reg.



## Summary statistics

What is the average delay on the departure of these flights?

Most of the functions we use for these operations comes from the library
dplyr, which is part of the tidyverse package.

The function to get summary statistics (average/mean, median, standard deviation etc)
from dataframes is *summarise*. It works like this:

```{r}
flightdata %>% 
  summarise(avg_dep_delay = mean(dep_delay))
```

Oops! That was not the result we wanted. 

## Missing data

Missing data is data that should be there. But isn't. The delay might have been
0 minutes. Or -5 minutes if the flight arrived early. But we should have a 
value for delay. There are many ways of handling missing data. In R, only one of
them is the right.

Missing data is encoded as the special value "NA". It means that we have no
idea what the value is. But that it should be there. It can be anything.

Therefore we cant calculate a mean value. The average of "4" and "no idea" is
"no idea", or, NA.

Most functions have an optional argument that specifies how missing values are
handled. 

We can find them in the help for that function, and for mean() it is specified
with "na.rm". By default "na.rm = FALSE", that is, the function will not remove
NA-values. 

If we set "na.rm = TRUE" missing values will be stripped from the data, before 
the average is calculated.

## once more, without NAs

```{r}
flightdata %>% 
  summarise(avg_dep_delay = mean(dep_delay, na.rm = T))
```

Note that we can write "T" instead of "TRUE". Likewise we could write "F" 
instead of "FALSE", saving 3-4 keystrokes.

### What else?
Other statistical parameters we might want to look at could be:

median
standard deviation
korrelationen, covarians etc.

Note that some of these functions handles missing values in different ways. 
Typically we add an argument na.rm = T to remove values, some of these functions
instead have an argument use, that can specify the way NAs are handled, eg:
```{r}
flightdata %>% 
  select(arr_delay, dep_delay) %>% 
  summarise(test = cor(arr_delay, dep_delay, use = "complete.obs"))
```


## Which airline is most on time?

If we want to get the averege departure delay for a specific company, we
could filter for a specific name of the carrier:

```{r}
flightdata %>% 
  filter(carrier == "UA") %>% 
  summarise(avg_dep_delay = mean(dep_delay, na.rm = T))
```

We could then do the same for every carrier. That is cumbersome.

Instead dplyr have a function that allow us to group the dataframe based on a
column, and do the subsequent operations on those groups:


```{r}
flightdata %>% 
  group_by(carrier) %>% 
  summarise(avg_dep_delay = mean(dep_delay, na.rm = T))
```

Much better! 

A second step would be to sort the carriers by the average departure delay:
```{r}
flightdata %>% 
  group_by(carrier) %>% 
  summarise(avg_dep_delay = mean(dep_delay, na.rm = T)) %>% 
  arrange(avg_dep_delay)
```

The carrier "US" does best. Who is worst?

*arrange* sorts, by default in ascending order. If we want it in descending order,
we can use a small helper function:

```{r}
flightdata %>% 
  group_by(carrier) %>% 
  summarise(avg_dep_delay = mean(dep_delay, na.rm = T)) %>% 
  arrange(desc(avg_dep_delay))
```


### summarising on more that one thing

The summarise function can summarise on more than one thing:

```{r}
flightdata %>%
  group_by(carrier) %>% 
  summarise(avg_dep_delay = mean(dep_delay, na.rm =T),
            avg_arr_delay = mean(arr_delay, na.rm=T)) 
```

Would this work?
```{r eval = F}
flightdata %>%
  group_by(carrier) %>% 
  summarise(avg_dep_delay = mean(dep_delay, na.rm =T)) %>% 
  summarise(avg_arr_delay = mean(arr_delay, na.rm=T)) 
```

why not?

{% include links.md %}
