---
title: "Joining data"
keypoints:
- FIXME
objectives:
- FIXME

questions:
- FIXME
teaching: 20
exercises: 10
source: Rmd
---


```{r, include = FALSE}
source("../bin/chunk-options.R")
knitr_fig_path("03-")
source("../bin/download_data.R")
library(dplyr)
library(readxl)
flightdata <- read_excel("../data/flightdata.xlsx", na = "NULL")
delays <- flightdata %>% 
  group_by(carrier) %>% 
  summarise(avg_dep_delay = mean(dep_delay, na.rm =T),
            avg_arr_delay = mean(arr_delay, na.rm=T))
  
```

## What is the actual name of those airlines?

It is nice to be able to identify the airline that is most on time. 
But we have carrier codes, not the actual names of them.

How do we get that?

### Reading another sheet in a spreadsheet

Our spreadsheet contains more that one sheet. The second sheet 

```{r eval = T}
carriers <- read_excel("../data/flightdata.xlsx", 2)
```

```{r eval = F}
carriers <- read_excel("data/flightdata.xlsx", 2)
```

```{r}
carriers
```

We now have a second data frame, containing the names of the carriers. And we have
a data frame containing the average delays

```{r}
delays
```

what we would like is something like this:
```{r echo = F, message = F, warning=F}
delays %>% left_join(carriers) %>% relocate(name, .after = carrier)
```

What we want to do is joining the two dataframes, so we enrich the original
delays dataframe with the name matching the carrier code.

There are 

## Can we plot the average delays?

Yes we can!

```{r}
delays %>%  ggplot(mapping = aes(x = avg_dep_delay, y = avg_arr_delay)) +
  geom_point()
  
```
Maybe it would be a good idea to color the points based on carrier. That is,
we would like to map the carrier to the color of the points:

```{r}
delays %>%  ggplot(mapping = aes(x = avg_dep_delay, y = avg_arr_delay, color = carrier)) +
  geom_point()
  
```

those points are not that useful
```{r}
delays %>%  ggplot(mapping = aes(x = carrier, y = avg_arr_delay, fill = carrier)) +
  geom_col() +
  coord_flip()
```


{% include links.md %}
