---
title: "boxplot og lineær regression (måske)"
teaching: 80
exercises: 35
questions:
  - "FIXME"
objectives:
  - "FIXME"
keypoints:
  - "FIXME"
   
source: Rmd
---

Endnu tættere blik på data - boxplots.



facetter

density curves - er der forskelle på de tre lufthavne?

Er der forskel på forsinkelser som funktion af flyenes alder?

lineær reg.


```{r, include=FALSE}
source("../bin/chunk-options.R")
knitr_fig_path("04-")
source("../bin/download_data.R")
library(tidyverse)
library(readxl)
flightdata <- read_excel("../data/flightdata.xlsx")
carriers <- read_excel("data/flightdata.xlsx", sheet = 2)

```


## Boxplots 

Boxplots is a common way of visualizing the most common summary statistics:


```{r}
flightdata %>% 
  group_by(carrier) %>% 
  summarise(avg_dep_delay = mean(dep_delay, na.rm = T),
            med_dep_delay = median(dep_delay, na.rm = T),
            sd_dep_delay = sd(dep_delay, na.rm = T),
            max_dep_delay = max(dep_delay, na.rm = T),
            min_dep_delay = min(dep_delay, na.rm  =T),
            iqr = IQR(dep_delay, na.rm = T)) %>% 
  arrange(avg_dep_delay) %>% 
  left_join(carriers) %>% 
  relocate(name, .after = carrier)

```

For this visualization, we use air_time, the time the flight takes, instead of
departure delays. 

```{r}
flightdata %>% 
  left_join(carriers) %>% relocate(name, .after = carrier) %>% 
  ggplot(mapping = aes(x = name, y = air_time)) + 
  geom_boxplot() + 
  coord_flip()
```

The boxplots show the inter quartile range with the white box. The solid black
line within that, is the median of the air_time. The horizontal lines, called
whiskers, on each
side of the white box shows the "minimum" and "maximum" observations, defined
as the observations that lies no longer from the IQR than 1.5 of the IQR. 

The solid dots at each end of the boxplot, represents outliers, observations that 
lies outside the whiskers.

Depending on the data, and the nature of the analyses we are going to do,
outliers are either very interesting, or something that we can ignore.

In this case it might be interesting to investigate which destinations Hawaiian
Airlines serve. Their fligts are much longer than the flights of the rest of 
the airlines. 

Let us take a look:
```{r}
flightdata %>% 
  left_join(carriers) %>% relocate(name, .after = carrier) %>% 
  filter(name == "Hawaiian Airlines Inc.") %>% 
  count(dest)
```
The `count()` function counts the number of times a value exists in a given 
column, in this case `dest`. 

Conclusion: Hawaiian Airlines only flies to Honolulu. 

Which airlines flies to Honolulu?

```{r}
flightdata %>% 
  left_join(carriers) %>% relocate(name, .after = carrier) %>%
  filter(dest == "HNL") %>% 
  count(name)
```
Here we filter in order to only have flights to Honolulu, and 
then count the different airlines. Only United Air Lines and Hawaiian 
Airlines have fligths to Honolulu.


## Boxplots are not necessarily a good idea

The boxplots of the air_time indicates that there is more structure in the 
data than boxplots can reveal. Maybe we should look at the distribution of
airtime amongst the airlines.

We could plot histograms of the airtime for each airline. Or we could plot
the smoothed histograms we saw in the correllograms for each airline. Or
we can use an extra package, `ggridges` to do something fancy.

Begin by installing the package:
```{r eval = F }
install.packages("ggridges")
```

Then load the package, and plot the data:
```{r}
library(ggridges)

flightdata %>% 
  left_join(carriers) %>% relocate(name, .after = carrier) %>% 
ggplot(aes(x = air_time, y = name, fill = name)) +
  geom_density_ridges() 
```


## What is the HVAD POKKER KALDER VI DET?

We saw earlier, that there was a correlation between departure delay and 
arrival delay. Can we use the data to predict how delayed we will be upon
arrival, when we know that the departure was 47 minutes delayed.

First, the plot:
```{r}
flightdata %>%
  sample_frac(0.005) %>% 
  ggplot(mapping = aes(x = dep_delay, y = arr_delay)) +
  geom_point() 
```
This looks more or less linear. We can place a linear regression line in the plot
using the function `geom_smooth(method = "lm")`, where we specify that the function
should fit a linear line to the data.

```{r}
flightdata %>%
  sample_frac(0.005) %>% 
  ggplot(mapping = aes(x = dep_delay, y = arr_delay)) +
  geom_point() +
  geom_smooth(method = "lm")
```
So, what is the actual linear model of this data?

A linear model in this case would be a formula on the form:

$arr_delay = a dep_delay + b$

The task is now to find the values of `a` and `b` that best fit the data. This is
called "fitting a linear model". 

Fitting a linear model to data, is done using the `lm()` function.

We need to specify the actual model that we want to fit, using the special
formula notation in R:

arr_delay ~ dep_delay

In addition we specify the input data in the function:

```{r}
model <- lm(arr_delay ~ dep_delay, data = flightdata)
```

We do this on the entire dataset.

Let us take a look at the result:
```{r}
model
```
We will not spend time on a complete discussion on p-values, but the way to get
them is to call the function `summary()` on the model:

```{r}
summary(model)
```



{% include links.md %}
